---
title: "Data Exploration: MSK-CHORD and MSK-MET"
author: "Maurin Widmer"
format: html
editor: visual
---

# Intent

In this exploratory data analysis I intend to get a feel for the two data sets MSK-CHORD and MSK-MET. I wont look in to a lot of details, but try and see in what ways the data is similar and in what ways they differ.

# Libraries

```{r}
library(tidyverse)
library(janitor)
library(skimr)
library(readr)
```

# MSK-CHORD

Targeted sequencing of 25040 tumors from 24950 patients and their matched normals via MSK-IMPACT, along with clinical annotations, some of which are derived from natural language processing (denoted NLP).

## 1. Loading and Overview

```{r}
## Load Data
msk_chord <- read_tsv("../Data/msk_chord_2024_clinical_data.tsv")
msk_chord   <- msk_chord   |> clean_names()
## Inspect Structure
glimpse(msk_chord)
```

## 2. Missing Data

```{r}
msk_chord |> 
  summarise(across(everything(), ~ sum(is.na(.)))) |> 
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  arrange(desc(n_missing))
```

## 3. Duplicates

```{r}
##looking for duplicates
sum(duplicated(msk_chord))
msk_chord |> 
  count(sample_id) |> 
  filter(n > 1)
```

## 4. Range and Validity Check

```{r}
summary(msk_chord)
```

# MSK-MET

Pan-cancer cohort of tumor genomic and clinical outcome data from 25,000 patients. The dataset identifies associations between tumor genomic alterations and patterns of metastatic dissemination across 50 tumor types; showing that chromosomal instability is strongly correlated with metastatic burden in some tumor types, like prostate and lung adenocarcinomas and HR+/HER2+ breast ductal carcinoma, but not in others, such as colorectal MSS, pancreatic adenocarcinoma and high-grade serous ovarian cancer. The study also identifies somatic alterations associated with increased metastatic burden and routes of metastatic spread. Our data offers a resource for the investigation of the biologic basis for metastatic spread and highlights the role of chromosomal instability in cancer progression.

## 1. Loading and Overview

```{r}
## Load Data
msk_met <- read_tsv("../Data/msk_met_2021_clinical_data.tsv")
msk_met   <- msk_met   |> clean_names()
## Inspect Structure
glimpse(msk_met)
```

## 2. Missing Data

```{r}
msk_met |> 
  summarise(across(everything(), ~ sum(is.na(.)))) |> 
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  arrange(desc(n_missing))
```

## 3. Duplicate

```{r}
## Searching for Duplicates
sum(duplicated(msk_met))
msk_met |> 
  count(sample_id) |> 
  filter(n > 1)
```

## 4. Range and Validity Check

```{r}
summary(msk_met)
```

# Checking if there is any overlap in data

## 1. Patient level

```{r}
colnames(msk_met)
colnames(msk_chord)
```

```{r}
## Looking at the number of common patient id
common_patients <- intersect(
  unique(msk_met$patient_id),
  unique(msk_chord$patient_id)
)

length(common_patients)

## Proportion of common patients
length(common_patients) / length(unique(msk_met$patient_id))
length(common_patients) / length(unique(msk_chord$patient_id))
```

We get that about half the patients contributed to both of the studies. (50.5% and 52.2%)

## 2. Sample level

```{r}
common_samples <- intersect(
  unique(msk_met$sample_id),
  unique(msk_chord$sample_id)
)

length(common_samples)

## Proportion of common patients
length(common_samples) / length(unique(msk_met$sample_id))
length(common_samples) / length(unique(msk_chord$sample_id))
```

And just under half the samples are used in both studies. (47,1%/48,5%)

## 3. Overlapping Sample Belong to Overlapping Patients

```{r}
## Checking if IDs are consistent
overlap_check <- msk_met |>
  filter(sample_id %in% common_samples) |>
  select(patient_id, sample_id) |>
  inner_join(
    msk_chord |> select(patient_id, sample_id),
    by = "sample_id",
    suffix = c("_met", "_chord")
  )

sum(overlap_check$patient_id_met != overlap_check$patient_id_chord)
```

Sample IDs are consistently mapped to the same patient in both data sets. There are no IDs with no mapping.

## 4. Checking if sample_id is consistent

Let's have a look at the patients that intersect and check if the patient data is the same to make sure they are really the same.

```{r}
# Extract all shared sample ids
shared_samples <- intersect(
  msk_chord$sample_id,
  msk_met$sample_id
)

# Only looking at the shared data to compare the metadata. This way we're shure that the campels are actually the same.
msk_chord_shared <- msk_chord |> 
  filter(sample_id %in% shared_samples)

msk_met_shared <- msk_met |> 
  filter(sample_id %in% shared_samples)

# Extract the column names both have in common
common_columns <- intersect(
  colnames(msk_chord),
  colnames(msk_met)
)

# 
msk_chord_shared <- msk_chord_shared |> select(all_of(common_columns))
msk_met_shared   <- msk_met_shared   |> select(all_of(common_columns))

# Create a dataframe that shows if a certein entry is equal in both data sets or not on NA
msk_shared_information <- msk_chord_shared == msk_met_shared
```

After inspecting the data frame it seems like the difference between the rows sharing the same sample ID are only differences in naming conventions (lymph vs. lymph node, neck vs. head neck) or small mistakes (unknown vs. female).

Next, I would like to check which tumor types are in the over lapping data

```{r}
unique(msk_chord_shared$cancer_type)
```

As expected, the cancer types in the MSK-CHORD data set are a strict subset of the MSK-MET data set. (attention not the data! only the cancer type)

# Which primary tumor tends to metastasize to the CNS

next we are searching in the MSK-MET data set for all the samples of metastatic tumors that lie in the CNS/Brain and then see if we find primary tumors that are more likely to spread to the CNS then others.

We start with the count:

```{r}
# get the counts of metastasis samples per primary tumor site. I included multiple samples per patient since they are separate metastasis events, but if you ant to exclude them just uncomment below
cns_met_counts <- msk_met |>
  filter(metastatic_site == "CNS/Brain") |>
  #distinct(patient_id, primary_tumor_site) |>
  count(primary_tumor_site, sort = TRUE)

cns_met_counts
```

Qualitatively we already see that breast cancer is the second most common primary tumor to CNS metastasis

As a second step, we have a look at the metastasis rate:

```{r}
cns_rate <- msk_met |>
  # distinct(patient_id, primary_tumor_site, metastatic_site) |>
  mutate(cns = metastatic_site == "CNS/Brain") |>
  group_by(primary_tumor_site) |>
  summarise(
    total_patients = n_distinct(patient_id),
    cns_patients = n_distinct(patient_id[cns]),
    cns_rate = cns_patients / total_patients
  ) |>
  arrange(desc(cns_rate))

cns_rate
```

Here we see that the rate of breast cancer to the CNS is at 0.016 (43 in 2612 cases).

## **Breast vs. all other tumors (Fisher’s exact test)**

Tests whether the proportion of CNS metastases among breast tumors differs from the proportion among all other tumors. A significant result means breast tumors are statistically over- or under-represented among CNS metastases compared with other cancers.

```{r}
patient_data <- msk_met |>
  #distinct(patient_id, primary_tumor_site, metastatic_site) |>
  mutate(cns = metastatic_site == "CNS/Brain",
    breast = primary_tumor_site == "Breast")

# creating a contingancy table of breast tumours and if they spread to the CNS or not
table_data <- table(patient_data$breast, patient_data$cns)

# Chi^2 test
fisher.test(table_data)
```

p value is not significant and null hypothesis can't be rejected.

## **Top ten tumor types (Chi-square test)**

Tests whether CNS metastasis frequency depends on tumor type across the 10 major tumor categories. A significant result means at least one tumor type has a different CNS metastasis rate than expected if tumor type and CNS spread were independent.

```{r}
top_tumors <- msk_met |>
  count(primary_tumor_site, sort = TRUE) |>
  slice_head(n = 10) |>
  pull(primary_tumor_site)

filtered <- patient_data |>
  mutate(
    tumor_group = ifelse(primary_tumor_site %in% top_tumors,
                         primary_tumor_site,
                         "Other")
  )

tab <- table(filtered$tumor_group, filtered$cns)

chisq.test(tab)
```

Here we just see that a sample (probably lung metastasis) seems to be significant.

## **Logistic regression (CNS metastasis \~ breast)**

Models the probability of a metastasis being in the CNS as a function of tumor type and estimatesodds ratios for each tumor type relative to a reference. A significant coefficient means that tumor type has significantly higher or lower odds of CNS metastasis compared with the reference tumor type.

```{r}
model <- glm(
  cns ~ breast,
  data = patient_data,
  family = binomial()
)

summary(model)
```

The breastTRUE coefficient is not significant. So if the primary tumor was a located in the breast is not a good predictor for metastasis to the CNS.

# Details of mutations and tumor types

In this section, I incorporate some of the next steps suggested by Kevin before moving on to the first tests with the model.

First we look at the number of primary tumors (and secondary) in in the CNS and breast.

```{r}
# number of primary and metastatic tumors (total)
msk_met |>
  count(sample_type)

# number of primary and metastatic breast tumors
msk_met |>
  filter(primary_tumor_site == "Breast") |>
  count(sample_type)

# number of primary and metastatic tumors by tumor type
msk_met |>
  count(cancer_type, sample_type)

# number of metastatinc breast tumor HER2+/her2
breast_data <- msk_met |>
  filter(cancer_type == "Breast Cancer") |>
  count(subtype, sample_type)
breast_data

breast_table <- table(
  msk_met$subtype[msk_met$cancer_type == "Breast Cancer"],
  msk_met$sample_type[msk_met$cancer_type == "Breast Cancer"]
)

chisq.test(breast_table)
```

A p-value of 0.2111 means we do not have statistical evidence that mutation type differs between primary and metastatic tumors. **-\> DISCUSS WITH KEVIN**

## Which Mutations are more common metastasis than in primary breast

First, we load the mutations data for msk_met and msk_chord

```{r}
# load mutations for MSK-MET
muts_met <- read.delim("../Data/msk_met_2021/data_mutations.txt")

# load mutations for MSK-CHORD
mut_chord <- read.delim("../Data/msk_chord_2024/data_mutations.txt")
```

Next, we attach sample information (Cancer_Type and sample_type) to the mutation information via the sample_id/Tumor_Sample_Barcode.

```{r}
# Attache sample meta data to mutations
muts_met_annot <- muts_met |>
  inner_join(
    msk_met |> select(sample_id, cancer_type, sample_type),
    by = c("Tumor_Sample_Barcode" = "sample_id")
  )

# Restrict to breast cancer
breast_muts <- muts_met_annot |>
  filter(cancer_type == "Breast Cancer")

# Count mutations per gene
gene_counts <- breast_muts |>
  count(Hugo_Symbol, sample_type, sort = TRUE)

gene_counts

# Count mutated samples per gene
gene_sample_counts <- breast_muts |>
  distinct(Tumor_Sample_Barcode, Hugo_Symbol, sample_type) |>
  count(Hugo_Symbol, sample_type, sort = TRUE)

# top genes in primary tumor
gene_sample_counts |>
  filter(sample_type == "Primary") |>
  slice_max(n, n = 10)

# top genes in metastases
gene_sample_counts |>
  filter(sample_type == "Metastasis") |>
  slice_max(n, n = 10)

# compute frequency for breast cancer
n_samples <- msk_met |>
  filter(cancer_type == "Breast Cancer") |>
  count(sample_type, name = "total_samples")

gene_sample_counts <- breast_muts |>
  distinct(Tumor_Sample_Barcode, Hugo_Symbol, sample_type) |>
  count(Hugo_Symbol, sample_type, name = "mutated_samples")

gene_freq <- gene_sample_counts |>
  left_join(n_samples, by = "sample_type")

gene_freq <- gene_freq |>
  mutate(freq = mutated_samples / total_samples)

gene_freq_wide <- gene_freq |>
  select(Hugo_Symbol, sample_type, freq) |>
  pivot_wider(
    names_from = sample_type,
    values_from = freq
  )

gene_freq_wide

# creating long dataset
gene_freq_long <- gene_freq_wide |>
  pivot_longer(
    cols = c("Primary", "Metastasis"),
    names_to = "Sample_Type",
    values_to = "Frequency"
  )

# prepare data for plotting
top_genes <- gene_freq_wide |>
  slice_max(Metastasis, n = 10) |>
  pull(Hugo_Symbol)

gene_freq_long_top <- gene_freq_long |>
  filter(Hugo_Symbol %in% top_genes)

# plotingt the top genes
ggplot(gene_freq_long_top, aes(x = Hugo_Symbol, y = Frequency, fill = Sample_Type)) +
  geom_col(position = "dodge") +
  labs(
    title = "Top 10 Breast Cancer Mutated Genes (Primary vs Metastasis)",
    x = "Gene",
    y = "Mutation Frequency"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

# CNS metastases originating from breast cancer vs. primary CNS tumors

We are now interested in the differences between primary CNS tumors and metastasis in the CNS that specifically originate from breast cancer. For that we define two groups:

**A. CNS metastases from breast cancer**

-   From `msk_met`:

    -   `sample_type == "Metastasis"`

    -   `primary_tumor_site == "Breast"`

    -   `metastatic_site == "CNS/Brain"`

**B. Primary CNS tumors**

-   From `tcga_data` (primary CNS data is not available in MSK):

    -   `sample_type == "Primary"`

    -   `cancer_type` = `"CNS/Brain"`

For Group B we take the TCGA data set because msk does not contain primary brain tumors. The data set is downloaded for a cohort of primary brain tumors. We start by loading the data:

```{r}
# Load TCGA primary brain tumors
tcga_data <- read_tsv("../Data/frequently-mutated-genes.2026-03-09.tsv")
```

First, we link the mutations with the two groups

```{r}
# CNS metastases from breast
cns_met_breast <- muts_met |>
  inner_join(msk_met |> 
               filter(sample_type == "Metastasis",
                      primary_tumor_site == "Breast",
                      metastatic_site == "CNS/Brain") |>
               select(sample_id, sample_type, primary_tumor_site, metastatic_site),
             by = c("Tumor_Sample_Barcode" = "sample_id"))

# Primary CNS tumors are ready from loaded data. We need to change the colnames 'symbol' = Hugo symbol and 'cohort_ssm_affected_cases_persfentage' = frequency
```

Next, we summarize mutation frequencies per gene

```{r}
# CNS metastases from breast mutation counts
cns_met_breast_counts <- cns_met_breast |>
  distinct(Tumor_Sample_Barcode, Hugo_Symbol) |>
  count(Hugo_Symbol, name = "n_samples") 

# Primary CNS
primary_cns_counts <- tcga_data |>
  rename(Hugo_Symbol = symbol, freq_tcga = cohort_ssm_affected_cases_percentage)

primary_cns_counts$freq_tcga <- primary_cns_counts$freq_tcga / 100 # percentage -> frequency

n_cns_met_breast <- length(unique(cns_met_breast$Tumor_Sample_Barcode))

cns_met_breast_counts <- cns_met_breast_counts |>
  mutate(freq = n_samples / n_cns_met_breast)

# Compare the two groups
combined_freq <- full_join(
  cns_met_breast_counts |> 
    rename(freq_msk = freq) |>      # make sure the column is called freq_msk
    select(Hugo_Symbol, freq_msk),
  
  primary_cns_counts |>
    select(Hugo_Symbol, freq_tcga),
  
  by = "Hugo_Symbol"
) |>
  replace_na(list(freq_msk = 0, freq_tcga = 0))
```

## Short visualization

```{r}
combined_long <- combined_freq |>
  pivot_longer(cols = c(freq_msk, freq_tcga),
               names_to = "Cohort",
               values_to = "Frequency") |>
  mutate(Cohort = recode(Cohort,
                         freq_msk = "CNS Metastasis (Breast)",
                         freq_tcga = "Primary CNS"))

# Top 10 genes in CNS metastases
top_genes <- combined_freq |>
  slice_max(freq_msk, n = 10) |>
  pull(Hugo_Symbol)

combined_long_top <- combined_long |>
  filter(Hugo_Symbol %in% top_genes)

ggplot(combined_long_top, aes(x = Hugo_Symbol, y = Frequency, fill = Cohort)) +
  geom_col(position = "dodge") +
  labs(
    title = "Mutation Frequencies in CNS Breast Metastases vs Primary CNS Tumors",
    x = "Gene",
    y = "Mutation Frequency"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Statistical test (Fisher's exact test)

For each mutation we perform a Fisher's exact test to see it there is a significant difference between primary CNS cancer and those metastasizing from breast cancer.

We prepare the count values for the

```{r}
# we have already prepared the count values for the metastasis
glimpse(cns_met_breast_counts)

# Prepare the tcga primary cns tumor count values
cns_primary_counts <- tcga_data |> 
  select(
    Hugo_Symbol = symbol,
    n_samples = num_cohort_ssm_affected_cases,
    freq = cohort_ssm_affected_cases_percentage
  )
cns_primary_counts$freq <- cns_primary_counts$freq / 100 # percentage -> frequency

glimpse(cns_primary_counts)
```

```{r}
# merging the cont data from primary and metstasis into one data set
cns_all_counts <- full_join(
  cns_met_breast_counts |> rename(n_msk = n_samples, freq_msk = freq),
  cns_primary_counts |> rename(n_tcga = n_samples, freq_tcga = freq),
  by = "Hugo_Symbol"
) |>
  replace_na(list(n_msk = 0, n_tcga = 0, freq_msk = 0, freq_tcga = 0))
#### IS IT REASONABLE TO DEAL WITH NA LIKE THIS? -> ASK KEVIN
```

```{r}
# MSK CNS breast metastases
total_msk <- length(unique(cns_met_breast$Tumor_Sample_Barcode))

# TCGA CNS primary tumors
total_tcga <- unique(tcga_data$num_cohort_ssm_cases)[1]
```

```{r}
# building the contigency table and add the p value of the fisher's exact test to the data
gene_tests <- cns_all_counts |>
  rowwise() |>
  mutate(
    freq_diff = freq_msk - freq_tcga,
    p_value = fisher.test(matrix(
      c(
        n_msk,
        total_msk - n_msk,
        n_tcga,
        total_tcga - n_tcga
      ),
      ncol = 2
    ))$p.value
  ) |>
  ungroup() |>   # VERY IMPORTANT
  mutate(
    p_adj = p.adjust(p_value, method = "BH")
  )

# look at the siginficant mutations
gene_tests |>
  filter(p_adj < 0.05) |>
  arrange(p_adj)
```

Let's visualize this in a plot

```{r}
top <- gene_tests |>
  filter(p_adj < 0.05) |>
  arrange(p_adj) |>
  slice_head(n = 10)

ggplot(top, aes(x = reorder(Hugo_Symbol, freq_diff), y = freq_diff)) +
  geom_col() +
  coord_flip() +
  labs(
    x = "Gene",
    y = "Frequency difference (MSK - TCGA)",
    title = "Top mutation differences"
  ) +
  theme_bw()
```

# Difference in mutation patterns between primary breast cancer and CNS metastasis originating in the breast

Now, we'd like to have the same plot as we just had for the different frequencies in primary CNS cancer and metastasis originating in the breast but look at the mutations in primary breast cancer and compare them to the mutations in CNS meastasis.

```{r}
# all mutations of primary breast cancer
breast_primary <- muts_met |>
  inner_join(msk_met |> 
               filter(sample_type == "Primary",
                      primary_tumor_site == "Breast") |>
               select(sample_id, sample_type, primary_tumor_site, metastatic_site),
             by = c("Tumor_Sample_Barcode" = "sample_id"))

# count of distinct muations in primary breast cancer
breast_primary_counts <- breast_primary |>
  distinct(Tumor_Sample_Barcode, Hugo_Symbol) |>
  count(Hugo_Symbol, name = "n_samples") 

# adding frequency of the metastasis
n_breast_primary <- length(unique(breast_primary$Tumor_Sample_Barcode))

breast_primary_counts <- breast_primary_counts |>
  mutate(freq = n_samples / n_breast_primary)
```

Next, we crate a data frame containing the counts and frequencies of all mutations in both primary breast cancers and its CNS metastasis and build a contingency table for each mutated gene and calculation the p value of fisher's exact test

```{r}
breast_all_counts <- full_join(
  cns_met_breast_counts |> rename(n_met = n_samples, freq_met = freq),
  breast_primary_counts |> rename(n_prim = n_samples, freq_prim = freq),
  by = "Hugo_Symbol"
) |>
  replace_na(list(n_met = 0, n_prim = 0, freq_met = 0, freq_prim = 0))

# total number of MSK primary breast cancer
total_prim <- n_breast_primary

# total number of MSK CNS metastasis with origin in the breast
total_met <- length(unique(cns_met_breast$Tumor_Sample_Barcode))

# building the contigency table and add the p value of the fisher's exact test to the data
gene_tests_breast <- breast_all_counts |>
  rowwise() |>
  mutate(
    freq_diff = freq_met - freq_prim,
    p_value = fisher.test(matrix(
      c(
        n_met,
        total_met - n_met,
        n_prim,
        total_prim - n_prim
      ),
      ncol = 2
    ))$p.value
  ) |>
  ungroup() |>
  mutate(
    p_adj = p.adjust(p_value, method = "BH")
  )
```

Now, we visualize just like we did before

```{r}
# searching for the top ten significt differences in mutations 
top_breast <- gene_tests_breast |>
  filter(p_adj < 0.05) |>
  arrange(p_adj) |>
  slice_head(n = 10)

# unfortunatelly non are significant so we procede with just the to performing
top_breast_unsig <- gene_tests_breast |>
  arrange(p_adj) |>
  slice_head(n = 10)

ggplot(top_breast_unsig, aes(x = reorder(Hugo_Symbol, freq_diff), y = freq_diff)) +
  geom_col() +
  coord_flip() +
  labs(
    x = "Gene",
    y = "Frequency difference (MET - PRIM)",
    title = "Top mutation differences"
  ) +
  theme_bw()
```

# Average number of Mutations

Here we calculate the average number of mutations in primary breast cancer, primary CNS cancer and metastasis in CNS originating from a breast tumor.

```{r}
# average number of mutations in primary breast cancer
avg_mut_breast <- muts_met |>
  filter(Tumor_Sample_Barcode %in% breast_primary$Tumor_Sample_Barcode) |>
  group_by(Tumor_Sample_Barcode) |>
  summarise(n_mut = n()) |>   # count number of mutations per sample
  summarise(avg_mut = mean(n_mut))

avg_mut_breast

# average number of mutations in primary CNS cancer
# however, the tcga data is not sample level -> we approx

# total mutations in the TCGA primary CNS cohort
total_mut_cns_tcga <- sum(tcga_data$num_cohort_ssm_affected_cases, na.rm = TRUE)

# number of samples in the cohort
n_samples_cns_tcga <- max(tcga_data$num_cohort_ssm_cases, na.rm = TRUE)

avg_mut_cns_primary <- total_mut_cns_tcga / n_samples_cns_tcga
avg_mut_cns_primary

#average number of mutations in metasasis in the CNS spread from breast
avg_mut_cns_met <- muts_met |>
  filter(Tumor_Sample_Barcode %in% cns_met_breast$Tumor_Sample_Barcode) |>
  group_by(Tumor_Sample_Barcode) |>
  summarise(n_mut = n()) |>
  summarise(avg_mut = mean(n_mut))

avg_mut_cns_met
```

```{r}
# combining it in a table
avg_mut_table <- tibble(
  cohort = c("Primary Breast", "CNS Met from Breast", "Primary CNS"),
  avg_mut = c(avg_mut_breast$avg_mut, avg_mut_cns_met$avg_mut, avg_mut_cns_primary)
)

avg_mut_table

#plotting
ggplot(avg_mut_table, aes(x = cohort, y = avg_mut, fill = cohort)) +
  geom_col() +
  theme_bw() +
  labs(y = "Average number of mutations per sample", x = "")
```
